---
title: "course project 1 - Analizing activity monitoring data"
author: "Federico Segui"
date: "February 9, 2018"
output: html_document
---

## Data
The dataset (Activity monitoring data) for this project was downloaded from this link:
https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip

#### The variables included in this dataset are:

* **steps**: Number of steps taking in a 5-minute interval (missing values are coded as )
* **date**: The date on which the measurement was taken in YYYY-MM-DD format
* **interval**: Identifier for the 5-minute interval in which measurement was taken

The dataset is stored in a comma-separated-value (CSV) file and there are a total of 17,568 observations in this dataset.

## Loading and preprocessing the data

```{r echo=TRUE}
tmp <- tempfile()
download.file("https://d396qusza40orc.cloudfront.net/repdata%2Fdata%2Factivity.zip", tmp)
unzip(tmp)
unlink(tmp)
data_activity <- read.csv("activity.csv")
data_activity[,2] <- as.Date(data_activity$date)
```

## What is mean total number of steps taken per day?

1. Calculate the total number of steps taken per day

2. Make a histogram of the total number of steps taken each day

3. Calculate and report the mean and median of the total number of steps taken per day:

```{r echo=TRUE}
options(digits = 10)
tot_steps <- tapply(data_activity$steps, data_activity$date, sum, na.rm=TRUE)
hist(tot_steps, xlab = "Total Steps per Day", ylab = "Frequency", main = "Total Steps per Day", col = "red")
steps_mean <-mean(tot_steps)
steps_median <- median(tot_steps)
```
The `mean` is **`r round(steps_mean,digits=2)`**

The `median` is **`r steps_median`**


## What is the average daily activity pattern?

1. Make a time series plot...
2. Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r echo=TRUE}
mean_steps <- aggregate(steps ~ interval, data_activity, mean)
plot(mean_steps$interval, mean_steps$steps, xlab = "Intervals", ylab = "Average steps", main = "Average No. of steps per day by interval", type = "l")
max_interval <- mean_steps[which.max(mean_steps$steps),1]
```
The 5-minute interval is **`r max_interval`**.

## Imputing missing values

1. Calculate and report the total number of missing values in the dataset (i.e. the total number of rows with s)

2. Devise a strategy for filling in all of the missing values in the dataset. The strategy does not need to be sophisticated. For example, you could use the mean/median for that day, or the mean for that 5-minute interval, etc.

3. Create a new dataset that is equal to the original dataset but with the missing data filled in.

4.Make a histogram of the total number of steps taken each day and Calculate and report the mean and median total number of steps taken per day. Do these values differ from the estimates from the first part of the assignment? What is the impact of imputing missing data on the estimates of the total daily number of steps?

```{r echo=TRUE}
missing_data <- sum(is.na(data_activity$steps) | is.na(data_activity$date) | is.na(data_activity$interval))

imp_data <- transform(data_activity, steps = ifelse(is.na(data_activity$steps), mean_steps$steps[match(data_activity$interval, mean_steps$interval)], data_activity$steps))

steps_day_imp<-with(imp_data,tapply(steps,date,sum,na.rm=TRUE))

hist(steps_day_imp, col = "blue",main = "Total Steps per Day", xlab = "Total Steps per Day", ylab = "Frequency")

mean_steps_day <- mean(steps_day_imp)
median_steps_day <- median(steps_day_imp)
```
```{r echo=FALSE}
cat("Missing values are: ", missing_data)
cat("Mean of the total steps: ", round(mean_steps_day,digits=2))
cat("Median of the total steps: ", round(median_steps_day,digits=2))
```

#### The media and median of the total steps by day after imputation of missing values are equal and differ from de media and median before imputation.

## Are there differences in activity patterns between weekdays and weekends?

1. Create a new factor variable in the dataset with two levels - "weekday" and "weekend" indicating whether a given date is a weekday or weekend day.

2. Make a panel plot containing a time series plot (i.e. ) of the 5-minute interval (x-axis) and the average number of steps taken, averaged across all weekday days or weekend days (y-axis). See the README file in the GitHub repository to see an example of what this plot should look like using simulated data.

```{r echo=TRUE}
weekdays <- c("Monday", "Tuesday", "Wednesday", "Thursday", "Friday")

imp_data$dow = as.factor(ifelse(is.element(weekdays(as.Date(imp_data$date)),weekdays), "Weekday", "Weekend"))

steps_day_imp <- aggregate(steps ~ interval + dow, imp_data, mean)

library(lattice)

xyplot(steps_day_imp$steps ~ steps_day_imp$interval|steps_day_imp$dow, main="Average Number of Steps per Day by Interval",xlab="Interval", ylab="Number of Steps",layout=c(1,2), type="l")

```